Fly away Peter, come back Paul: the conservation and losses of avian non-coding RNAs

### 2013-10-22 paper/manuscript.tex
Making it look like an article, updated phylogenetic orderings.

The bulk of the ``unusually well conserved RNAs'' belong to the long
non-coding RNA (lncRNA) group.  The lncRNAs are a diverse group of
RNAs that have been implicated in a multitude of functional processes
\cite{Rinn:2007,Chow:2005,Guttman:2009}. These RNAs have largely been
characterised in mammalian species, particularly human and
mouse. Consequently, we generally do not expect these to be conserved
outside of mammalia. Notable examples include Xist \cite{Duret:2006}
and H19 \cite{Smits:2008}. Some of the main results of this group are
discussed in the comparative analysis of avian genomes article \cite{}.

@@

When analysing the RNA-domain annotations it is striking that many of
the lncRNAs with multiple RNA-domains are consistently preserved in
the birds. The annotations of these domains lie in the same genomic
region, in the same order as in the mammalian homologs. Thus
supporting a high degree of evolutionary conservation for the entire
lncRNA. In particular the HOXA11-AS1, PCA3, RMST, Six3os1, SOX2OT and
ST7-OT3 lncRNAs have multiple, well conserved RNA-domains (See
Figure~\ref{fig:5}).

@@

Our ability to find additional homologs for several RNA families that fill
gaps in the abundance matrices (Supplemental Figure~5)
strongly suggests that that conspicuous absences, in particular of LUCA and
LECA RNAs, are caused by incomplete data in the current assemblies and
sequence divergence rather then true losses.

\subsection*{Y RNA clusters}

@@

\subsection*{MicroRNA clusters}

@@

In this work we have summarised the main results after carefully
annotating non-coding RNAs in genome sequences. We have identified
previously unrecognised conservation of ncRNAs in avian genomes as
well as some surprising ``losses'' of otherwise well conserved
ncRNAs. We suspect many of these losses are due to a combination of
limitations in the homology search tools that we use for annotation
and the ability of ncRNAs to tolerate large amounts of sequence
variation while remaining functional, rather than \emph{bona fide}
gene loss. In some cases these losses could be due to missing data
from the genome assemblies, but this unlikely to be the case for
multiple independant assemblies.

These results indicate we are still in the very early phases of
determining the functions of many RNA families. This is illustrated by
the fact that the reported functions of some ncRNAs are
mammal-specific, yet some are also found in bird genomes.

@@

@@

\begin{figure}[ht]
  \includegraphics[width=0.45\textwidth]{figures/lncRNA.pdf}
  \includegraphics[width=0.1\textwidth]{figures/key2.pdf}
  \caption[]{Heatmaps showing the prescence/abscence and approximate
    copy number of {\bf lncRNA} families.}\label{fig:5}
\end{figure}

>>>>

The bulk of the ``unusually well conserved RNAs'' belong to the long
non-coding RNA (lncRNA) group.  The lncRNAs are a diverse group of
RNAs that have been implicated in a multitude of functional processes
\cite{Rinn:2007,Chow:2005,Guttman:2009}. These RNAs have largely been
characterised in mammalian species, particularly human and
mouse. Consequently, we generally do not expect these to be conserved
outside of mammalia. Notable examples include Xist \cite{Duret:2006}
and H19 \cite{Smits:2008}.  There is emerging evidence for the
conservation of “mammalian” lncRNAs in other vertebrates
\cite{Chodroff:2010,Ulitsky:2011}), however, like most lncRNAs, the
function of these lncRNAs remains largely unknown. Here, we show the
conservation of several well-characterised lncRNAs of known function
in humans.

@@

When analysing the RNA-domain annotations it is striking that many of
the lncRNAs with multiple RNA-domains are consistently preserved in
the birds. The annotations of these domains lie in the same genomic
region, in the same order as in the mammalian homologs. Thus
supporting a high degree of evolutionary conservation for the entire
lncRNA. In particular the HOXA11-AS1, PART1, PCA3, RMST, Six3os1, SOX2OT and
ST7-OT3 lncRNAs have multiple, well conserved RNA-domains (See
Figure~\ref{fig:5}).

 The conservation of these ``human'' lncRNAs among birds
suggests they may also be functional in birds but what these
functions is not immediately obvious. For example, PART1 and
PCA3 are both described as prostate-specific lncRNAs that play a role
in the human androgen-receptor pathway
\cite{Bussemakers:1999,Lin:2000,Ferreira:2012}. Birds lack a prostate
but both males and females express the androgen receptor (AR or NR3C4) in gonadal
and non- gonadal tissue
\cite{Yoshimura:1993,Veney:2004,Fuxjager:2012,Leska:2012}. Thus, we
postulate that PART1 and PCA3 also play a role in the
androgen-receptor pathway in birds but whether the expression of these
lncRNAs are tissue specific is unknown at present.

@@

Our ability to find additional homologs for several RNA families that fill
gaps in the abundance matrices (Figure~\ref{fig:7})
strongly suggests that that conspicuous absences, in particular of LUCA and
LECA RNAs, are caused by incomplete data in the current assemblies and
sequence divergence rather then true losses.

%\subsection*{Y RNA clusters}

@@

%\subsection*{MicroRNA clusters}

@@

In this work we have summarised the main results after carefully
annotating non-coding RNAs in genome sequences using homology-based
methods. The homology-based tools have distinct advantages over
experimental-based approaches as not all RNAs are expressed in any
particular tissue-type or developmental-stage, in fact some RNAs have
extremely specific expression profiles \cite{Johnston:2003}.

 We have identified previously unrecognised conservation of
ncRNAs in avian genomes as well as some surprising ``losses'' of
otherwise well conserved ncRNAs. We suspect many of these losses are
due to a combination of limitations in the homology search tools that
we use for annotation and the ability of ncRNAs to tolerate large
amounts of sequence variation while remaining functional, rather than
\emph{bona fide} gene loss. In some cases these losses could be due to
missing data from the genome assemblies, but this unlikely to be the
case for multiple independant assemblies.

These results indicate we are still in the very early phases of
determining the functions of many RNA families. This is illustrated by
the fact that the reported functions of some ncRNAs are
mammal-specific, yet some are also found in bird genomes. CITE DAVE
BURT ET AL AND THE COMPARATIVE ARTICLE?

@@

The snoRNA and miRNA annotations were validated using RNA-seq data
(\ppg{SOURCED FROM?}). Since not all RNAs are expressed at any given
time we expect to verify a fraction of RNAs annotated using expression
data. Never-the-less 242 of 691 (35\%) miRNAs validated as well as 328
of 376 (87\%) snoRNAs of our homology-based annotations were validated
using this data.

@@

\begin{figure}[ht]
  \includegraphics[width=0.45\textwidth]{figures/unusual-conserved.pdf}
  \includegraphics[width=0.45\textwidth]{figures/diverged.pdf}
  \includegraphics[width=0.1\textwidth]{figures/key2.pdf}
  \caption[]{Heatmaps showing the prescence/abscence and approximate
    genomic copy number of ``unusually, well conserved RNAs''
    (particularly the lncRNAs) on the left and families that have been
    identified as surprising RNA Losses, divergence or missing
    data. In several cases functionally related families have also
    been included, e.g. the RNA components of the major and minor
    spliceosomes: U1, U2, U4, U5 and U6; and U11, U12, U4atac, U5 and
    U6atac, respectively.}\label{fig:5}
\end{figure}

\begin{figure}[ht]
  \includegraphics[width=0.75\textwidth]{figures/More-cand.pdf}
  \caption[]{...}\label{fig:7}
\end{figure}
